---
title: "Lab_7_Assignment_Group_23"
format:
  html:
    embed-resources: true
editor: visual
---

# **Lab 7 Assignment: Group 23**

# Lab 7: Collaborative Bio Data Science using GitHub via RStudio

**Group 23, members:**

-   Ana Pastor Mediavilla - s222761

-   Carlos de Santiago León - s222766

-   Laura Figueiredo Tor - s222797

-   Monika Karolina Borkowska - s223203

-   Amanda Jimenez Lobato - s233150

## 1. Load libraries

```{r}
library('tidyverse')
library('broom')
library('patchwork')
library('ggrepel')

if(!require('cowplot')) {
  install.packages('cowplot')
  library('cowplot')
}
```

## 2. Load data

```{r}
#| message: false 
biopsy <- read_csv("https://wilkelab.org/classes/SDS348/data_sets/biopsy.csv")

biopsy
```

## 3. Analysis and Results

### 3.1. Data in PC coordinates

We fit first the PCA method according to the data by selecting the numeric features of the dataset and scaling them to have unit variance.

```{r}
pca_fit <- biopsy %>% 
  select(where(is.numeric)) %>% 
  prcomp(scale = TRUE)
```

```{r}
pca_fit <- biopsy %>% 
  select(where(is.numeric)) %>% 
  scale() %>% 
  prcomp()
```

Then, we reconstruct the data into the PC components and plot them, for analyzing possible relation/clustering according to the variable of interest (which in our study is the target/outcome related to the presence of a bening or malignant tumour).

```{r}
pca_fit %>%
  augment(biopsy) %>% 
  ggplot(aes(.fittedPC1, 
             .fittedPC2, 
             color = outcome)) + 
  geom_point(size = 1.5) +
  labs(title = "PCA scores",
       x = "Fitted PC1", 
       y = "Fitted PC2",
       color = "Outcome") + 
  scale_color_manual(values = c(malignant = "#D55E00", 
                                benign = "#0072B2")) +
  theme_half_open(12) + 
  background_grid() +
  theme(plot.title = element_text(hjust = 0.5))
```

From this PCA scores plot, we can observe a clear clustering according to the outcome variable by only looking at the two first principal components. This means that some numeric information from some of the data variables is shared between the two different outcomes. For example, benign data points can have similar values in variables like the uniform_cell_size, while the malignant data may have a different range.

This clear differential clustering between the two data labels suggests that the dataset could be applied for further studies and applications such as machine learning algorithms for cancer diagnosis.

### 3.2. Rotation matrix

We will proceed to extract the rotation matrix. The rotation matrix consist of a set of vectors which provide the direction of the principal components, therefore the rotation matrix contains the eigenvectors.

```{r}
pca_fit %>%
  tidy(matrix = "rotation")
```

The roration matrix resulting from a PCA can be visualized in order to understand better how the variables contribute to the principal components and how this components are related to each other. This can be done by looking at the direction and length of the plotted arrows.

```{r}
# Arrow style for plotting
arrow_style <- arrow(
  angle = 20,
  ends = "first",
  type = "closed",
  length = grid::unit(8, "pt")
)

# Plot the rotation matrix by plotting the loadings

pca_fit %>%
  tidy(matrix = "rotation") %>%
  pivot_wider(names_from = "PC",
              names_prefix = "PC",
              values_from = "value") %>%
  ggplot(aes(PC1,
             PC2)) +
  geom_segment(xend = 0,
               yend = 0, 
               arrow = arrow_style,
               color = 'darkblue') +
  geom_text_repel(aes(label = column),
                  box.padding = 0.5,
                  segment.color = "darkgreen",
                  size = 4) +

  xlim(-1.25, .5) +
  ylim(-.5, 1) +
  coord_fixed() + # fix aspect ratio to 1:1
  theme_minimal_grid() +
  theme(plot.title =  element_text(hjust = 0.5, size = 24, face = "bold")) +
  labs(title = "PCA Rotation Matrix Plot", x = "PC1", y = "PC2")
```

From this plot we can see how many of the variables are closely related to each other, as is the case of `bare_nuclei` , `clump_thickness`, and `bland_chromatin`, where we can see how the rows have almost the same length and are pointing in the same direction. On the other hand, these variables are not related to `mitoses`, as the arrows are almost perpendicular (orthogonal), suggesting that they are uncorrelated. The length of the arrows represents the variance explained by each principal component, where longer arrows imply that the corresponsing variable captures more variance of the data, this is the case with the variable `mitoses`, which captures a great amount of the variance. Further analysis on the variance explained by each component can be done in order to identify which components capture the most variety of data.

### 3.3. Variance

## 4. Discussion
